D3064R0: OOTA Execution on Real Implementations is Provably Vacuous

### 2024-01-06 ootavacuous.tex
Check and wordsmith non-JMM litmus tests
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

This appendix evaluates a number of OOTA-related litmus tests.
The evaluations are carried out using the \co{herd7} tool and also
using manual analysis.
The manual analysis uses the informal rule that a semantic dependency
exists between a given relaxed atomic load and store on any execution
where a different value loaded would change the address stored to, the
value stored, or the whether or not that store would be executed in
the first place.
Producing a precise formal statement of this rule turns out to be quite
challenging.
For but one example, identical stores are fungible, so that if all
executions always execute one of several equivalently positioned stores
having identical addresses and values, there is no semantic dependency
from any load to any of those stores~\cite{Howells2009membartxt}.

@@

Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4}
shows a different way of expanding
Listing~\ref{lst:Causality Test Case 4}
(discussed in
Appendix~\ref{app:Causality Test Case 4})
from two processes to three processes.
The \co{herd7} tool finds OOTA values, but in a rather odd manner.

@@

One possible reaction to this situation is to declare that useful
real-world C++ implementations should never duplicate non-volatile
relaxed atomic loads.
But suppose that the initial code was as shown in
Listing~\ref{lst:Unused Extra Atomic Load}?\footnote{
	Such unused loads can easily be generated by complex macros or
	template metaprograms.}
In this case, the transformation to
Listing~\ref{lst:Atomic Loads Invented}
would not require inventing a load, but instead merely making use of
the previously discarded value in \co{r0}.

>>>>

This appendix evaluates a number of OOTA-related litmus tests.
The evaluations are carried out using the \co{herd7} tool and also
using manual analysis.

The manual analysis uses the informal rule that a semantic dependency
exists between a set of relaxed atomic loads and a store on any execution
where different values loaded would change the address stored to, the
value stored, or the whether or not that store would be executed in
the first place, with the proviso that stores of the same value to the
same address having the same ordering properties are fungible.
Producing a precise formal statement of this rule turns out to be quite
challenging, but as noted earlier, there is considerable work along
various
lines~\cite{Lahav:2017:RSC:3062341.3062352,Sinclair:2017:CAR:3079856.3080206,Lee:10.1145/3385412.3386010,MarkBatty2019ModularRelaxedDependenciesOOTA}.
For examples of the challenges, please see
Section~\ref{sec:On OOTA Definitions}.

@@

Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4}
shows a different way of expanding
Listing~\ref{lst:Causality Test Case 4}
(discussed in
Appendix~\ref{app:Causality Test Case 4})
from two processes to three processes.
The \co{herd7} tool finds OOTA values, as expected.

@@

One possible reaction to this situation is to declare that useful
real-world C++ implementations should never duplicate non-volatile
relaxed atomic loads.
But suppose that the initial code was as shown in
Listing~\ref{lst:Unused Extra Atomic Load}?\footnote{
	Such unused loads can easily be generated by complex macros or
	template metaprograms.}
In this case, the transformation to
Listing~\ref{lst:Atomic Loads Invented}
would not require inventing a load, but instead merely inventing a use of
the previously discarded value in \co{r0}.
